An Application of Graph Theory to MDAO Problem Formulation

### 2013-02-21 graph_syntax.tex
update

\section{Graph-based Problem Formulation Syntax}
We begin by discussing the general syntax of graph theory and then expand upon this syntax to represent the data flow of an MDAO problem.
A graph is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, which means that the elements of $E$ are 2--element subsets of $V$ \cite{Diestel2010}. 
For a directed graph we introduce two maps $S:E \to V$ and $F:E \to V$ to denote the starting and final nodes of each edge, respectively. 
\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=1.5in]{images/example_directed_graph}
	\end{center}
	\vspace{-20pt}
\caption{Example directed graph.}
\label{f:example directed graph}
\end{figure}
As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
\begin{IEEEeqnarray*}{rCl}
V & = & \{1,2,3,4,5\}, \\
E & = & \big\{\{1,2\},\{2,3\},\{1,3\},\{3,4\}\big\}, \\
S & = & \Big\{ \big(\{1,2\},2\big),\big(\{2,3\},3\big),\big(\{1,3\},1\big),\big(\{3,4\},3\big) \Big\}, \\
F & = & \Big\{ \big(\{1,2\},1\big),\big(\{2,3\},2\big),\big(\{1,3\},3\big),\big(\{3,4\},4\big) \Big\}.
\end{IEEEeqnarray*}

We now assign additional properties to the nodes and edges so that they may represent the data flow of an MDAO problem. 
The first property assigned to nodes and edges is the \emph{type}. The possible node types are:
\begin{itemize}
\item global input
\item global output
\item local input
\item local output
\item analysis start
\item analysis finish
\end{itemize}
and the possible edge types are
\begin{itemize}
\item connection
\item in
\item analysis
\item out
\end{itemize}
Then, considering sets $T_\txt{node}$ and $T_\txt{edge}$ to be sets containing the possible node types and edge types, respectively, we use mappings $t_\txt{node}:V \to T_\txt{node}$ and $t_\txt{edge}:E \to T_\txt{edge}$ to assign a type to each node and edge. Nodes and edges are named according to their type (e.g. connection edge). Additional properties will be discussed is subsequent sections.

@@

Next, Consider an MDAO problem defined by a set of codes, global inputs, and global outputs, as represented by analysis blocks $A_i=(V_{A_i},E_{A_i}),i=1,\ldots,m$, the set of nodes $\mathcal I$, and the set of nodes $\mathcal O$, respectively. 
Then we may construct the \emph{maximal connectivity graph} $M=(V_M,E_M)$ as
\begin{IEEEeqnarray*}{rCl}
V_M & = & \mathcal I \cup \mathcal O \cup \left( \bigcup_{i = 1}^m V_{A_i} \right), \\
E_M & = & C \cup \left( \bigcup_{i=1}^m E_{A_i} \right),
\end{IEEEeqnarray*}
where $C$ is a set containing connection edges defined for $v_1,v_2 \in V_M$ as:
\begin{IEEEeqnarray*}{rl}
c = & \{v_1,v_2\} \in C \txt{ if:} \\
& t_\txt{node}(v_1) = \txt{`local output' or `global input'}, \\
& t_\txt{node}(v_2) = \txt{`local input' or `global output'},\txt{ and} \\
& n(v_1) = n(v_2),
\end{IEEEeqnarray*}
in which case $S(c) = v_1$ and $F(c) = v_2$. The maximal connectivity graph contains all possible connections that may exist between the given analysis blocks, and these connections are made by connecting variables with the same name, which requires a consistent naming convention between analysis blocks.

\subsection{Fundamental Problem Formulation Graph}
We may now define the graph representing the fundamental problem formulation $P=(V_P,E_P)$ as a subgraph of $M$ such that there are no \emph{holes} or \emph{collisions}. 
Before defining these terms, let us first define $\txt{deg}^-(v)$ as the \emph{indegree} $\txt{deg}^+(v)$ as the \emph{outdegree} of vertex $v$, where indegree and outdegree refer to the number of edges directed into vertex $v$ and the number of edges directed out of vertex $v$, respectively.
Let $H$ be the set of all nodes which are holes in $M$ and define $H$ as
\begin{equation}
v \in H \txt{ if } t_\txt{node}(v) = \txt{`local input' or `global output'} \txt{ and } \txt{deg}^-(v) = 0.
\end{equation}
Next, let $L$ be the set of all nodes which are collisions in $M$ and define $L$ as
\begin{equation}
v \in L \txt{ if } t_\txt{node}(v) = \txt{`local input' or `global output'} \txt{ and } \txt{deg}^-(v) > 1.
\end{equation}
This means that exactly one edge must be directed into each local input node and each global output node. We also require $P$ to contain all of the global output nodes, meaning $P\subset M$ is a fundamental problem formulation if
\begin{equation}
\mathcal O \subset V_P \ \txt{ and } \ v \notin H, v \notin L \ \forall v \in V_P.
\end{equation}

\subsection{Obtaining the Fundamental Problem Formulation Graph}
In general, the fundamental problem formulation graph $P$ may not exist, or there may be multiple different solutions. Then the first problem is to determine if there exists a graph $P$.
In practice, it is likely that this definition will not produce a unique solution. 
A collision represents a choice.

>>>>

\section{Graph-based Problem Formulation Syntax}
We begin by discussing the general syntax of graph theory and then expand upon this syntax to represent the data flow of an MDAO problem.
A graph is a pair $G = (V,E)$ of sets such that $E \subseteq V \times V$, which means that the elements of $E$ are 2--element subsets of $V$ \cite{Diestel2010}. 
For a directed graph we construct $E$ as a set of ordered pairs instead of a set of sets. Each ordered pair represents an edge starting at the node indicated by the first entry and directed to the node indicated by the second entry. Therefore, $E$ is a mapping from $V$ to $V$.
\begin{figure}[htb!]
	\begin{center}
	\includegraphics[width=1.5in]{images/example_directed_graph}
	\end{center}
	\vspace{-20pt}
\caption{Example directed graph.}
\label{f:example directed graph}
\end{figure}
As an example, for the directed graph shown in Fig.~\ref{f:example directed graph} we have
\begin{IEEEeqnarray*}{rCl}
V & = & \{1,2,3,4,5\}, \\
E & = & \big\{(1,2),(3,2),(1,3),(3,4)\big\}.
\end{IEEEeqnarray*}

We now assign additional properties to the nodes and edges so that they may represent the data flow of an MDAO problem. 
The first property assigned to nodes and edges is the \emph{type}. The possible node types are:
\begin{itemize}
\item global input
\item global output
\item local input
\item local output
\item analysis start
\item analysis finish
\end{itemize}
and the possible edge types are
\begin{itemize}
\item connection
\item in
\item analysis
\item out
\item ((consider eliminating `in' and `out' and using connection))
\end{itemize}
Then, considering sets $T_\txt{node}$ and $T_\txt{edge}$ to be sets containing the possible node types and edge types, respectively, we use mappings $t_\txt{node}:V \to T_\txt{node}$ and $t_\txt{edge}:E \to T_\txt{edge}$ to assign a type to each node and edge. Nodes and edges are named according to their type (e.g. connection edge). Additional properties will be discussed is subsequent sections.

@@

Next, Consider an MDAO problem defined by a set of codes, global inputs, and global outputs, as represented by analysis blocks $A_i=(V_{A_i},E_{A_i}),i=1,\ldots,m$, the set of nodes $\mathcal I$, and the set of nodes $\mathcal O$, respectively. 
Then we may construct the \emph{maximal connectivity graph} $M=(V_M,E_M)$ as
\begin{IEEEeqnarray*}{rCl}
V_M & = & \mathcal I \cup \mathcal O \cup \left( \bigcup_{i = 1}^m V_{A_i} \right), \\
E_M & = & C \cup \left( \bigcup_{i=1}^m E_{A_i} \right),
\end{IEEEeqnarray*}
where $C$ is a set containing connection edges defined for $v_1,v_2 \in V_M$ as:
\begin{IEEEeqnarray*}{rl}
c = & (v_1,v_2) \in C \txt{ if:} \\
& t_\txt{node}(v_1) = \txt{`local output' or `global input'}, \\
& t_\txt{node}(v_2) = \txt{`local input' or `global output'},\txt{ and} \\
& n(v_1) = n(v_2).
\end{IEEEeqnarray*}
The maximal connectivity graph contains all possible connections that may exist between the given analysis blocks, and these connections are made by connecting variables with the same name, which requires a consistent naming convention between analysis blocks.

\subsection{Fundamental Problem Formulation Graph}
Before defining the the fundamental problem formulation graph, let us first define $\txt{deg}^-(v)$ as the \emph{indegree} $\txt{deg}^+(v)$ as the \emph{outdegree} of vertex $v$, where indegree and outdegree refer to the number of edges directed into vertex $v$ and the number of edges directed out of vertex $v$, respectively. Then, define a mapping $d:V|_\txt{local input} \to \mathbb{N}$ to specifiy the maximum number of connections that are allowed into a local input node.
If $d(v)=1$ for each local input node, then exactly one edge must be directed into each local input node and each global output node. 
If $d(v)>1$ for a local input node, this means that multiple inputs are accepted, such as the case for a multifidelity formulation.

Next, let $H$ be the set of all nodes which are \emph{holes} in $M$ and define $H$ as
\begin{equation}
H=\big\{v \in M \big| t_\txt{node}(v) = \txt{`local input' or `global output'} \txt{ and } \txt{deg}^-(v) = 0\big\}.
\end{equation}
Also, let $L$ be the set of all nodes which are \emph{collisions} in $M$ and define $L$ as
\begin{equation}
L = \big\{v \in M \big| t_\txt{node}(v) = \txt{`local input' or `global output'} \txt{ and } \txt{deg}^-(v) > d(v) \big\}.
\end{equation}
We may now define the graph representing the fundamental problem formulation $F=(V_F,E_F)$ as a subgraph of $M$ such that there are no holes or collisions.

We also require $F$ to contain all of the global output nodes, meaning $F\subset M$ is a fundamental problem formulation if
\begin{equation}
\mathcal O \subset V_F \ \txt{ and } \ v \notin H, v \notin L \ \forall v \in V_F.
\end{equation}
((this is missing any edges!!!))
\subsection{Obtaining the Fundamental Problem Formulation Graph}
In general, the fundamental problem formulation graph $P$ may not exist, or there may be multiple different solutions. Then the first problem is to determine if there exists a graph $P$.
In practice, it is likely that this definition will not produce a unique solution. 
A collision represents a choice.
